\documentname: Is machine learning good or bad for the natural sciences?

### 2023-12-26 paper/main.tex
uses of ML in science

% To do:
% ------
% - Write an outline.
% - Fill in the outline.
% - Publish on the arXiv.
% - format for ICML Position Paper and submit.
% - Wait for the call from Stockholm.

@@

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
  For example, when a flexible machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
  We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when flexible models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when flexible regressions are used to make datasets, the elements of those datasets cannot be used in joint analyses without taking on uncontrolled biases.
  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
\end{abstract}

@@

\paragraph{Our contributions:}
\begin{itemize}
  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines. One is a confirmation bias that arises when simulations are replaced or augmented by emulators. Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}. Many of these places are in the operational parts of scientific projects.
  \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most flexible ML method delivers \emph{the most conservative approach} to the scientific problem. An example is in causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
\end{itemize}

@@

\section{Why do we need machine learning in the natural sciences?}
In this \sectionname{} we give some idea of how and why ML has had such a big impact in the natural sciences, and how and why many scientific projects require ML technologies.
In most cases, the ML is required on the engineering or execution side of projects that are large-scale in some sense.
\begin{description}
  \item[Label transfer:] Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
  In this case, if it is extremely expensive to label more objects, a regression can be trained on the labeled data points and then the trained regression used to label
  \item[Classification:]
  \item[Speeding up decisions:]
  \item[Speeding up simulations:]
  \item[Modeling nuisances:]
  \item[Outlier detection:]
  \item[Making discoveries?]
\end{description}
The short summary is that we need ML in the natural sciences; we can't live without it.
The question here is how we will use it safely?

@@

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases, and potentially produce unbiased estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace regressions with generative models that can produce likelihoods HOGG CITE THINGS.

@@

@@

\section{When is machine learning good for natural science?}\label{sec:good}

@@

\clearpage\appendix
\section{Population-level biases in regression outputs}\label{app:toy}
A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of a flexible function $f(x;\theta)$ from a training set of labeled data.
The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.

To illustrate these ideas, we construct a toy data analysis.
This toy will make a general point, but it has a decidedly astronomical feel to it.
The complete toy data set is generated by a process with the algorithm below.
This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
\begin{hoggnumerate}
    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
    \item The point is discarded if the age is outside the range $0<\eta<13$. If the point is not discarded:
    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$.
    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
\end{hoggnumerate}
The data generated this way are shown in the top panels of \figurename~\ref{fig:regression}.
The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
This relationship is shown in \figurename~\ref{fig:regression}.
It isn't precisely linear because of the censoring of data points with latent ages outside the range $0<\eta<13$.

We construct a three-layer multilayer perceptron (MLP) (using the scikit-learn implementation) with layer sizes of 64, 32, and 16 neurons.
The model is trained on a training set of 4096 data points with features $x$ and labels $y$.
The trained model takes as input features $x$ and outputs estimated labels $\hat{y}$.
The model is validated on a validation set of 2048 data points with features $x$ and labels $y$.
The test set is a much larger set of 105,864 data points with features $x$ and no labels.
The model is used to make estimated labels $\hat{y}$ in the full test set.
The validation and test labels are shown in the bottom panels of \figurename~\ref{fig:regression}.
Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
\begin{figure}[p!]
\includegraphics[width=0.49\textwidth]{notebooks/data_examples.png}
\includegraphics[width=0.49\textwidth]{notebooks/training_data.png} \\
\includegraphics[width=0.49\textwidth]{notebooks/validation.png}
\includegraphics[width=0.49\textwidth]{notebooks/test_data_results.png}
\caption{\sffamily%
Visualization of the toy regression. \textsl{Top-left:} Random examples of the data vectors $x$, which are one-dimensional images generated from a linear model plus a nonlinearity created by two rectifications. Details of the data generation are given in the text. Each example $x$ is labeled on the right side by the value of its label $y$. \textsl{Top-right:} The training-set labels $y$, plotted against the known parameter $r$, which is not used in the regression (only the vectors $x$ are used). Also shown are solid red circles showing the true mean relationship between $y$ and $r$ in the toy data. Open black squares show the empirical mean relationship measured in bins in the training-set data. \textsl{Bottom-left:} Validation of the trained regression in the validation set, showing that the label estimates $\hat{y}$ are noisy (as expected given the problem set-up) but not strongly biased. \textsl{Bottom-right:} The regression estimates $\hat{y}$ in the very large test set, plotted against the known parameter $r$. Also shown are the same solid red circles and open black squares as in the top-right plot. Black X-shaped symbols show the mean relationship between $\hat{y}$ and $r$. The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
\end{figure}

The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
They are not precisely estimated to be sure, but there is no evidence for strong bias at the individual-object level.

The true relationship between label $y$ (measured age) and known parameter $r$ (guiding radius) is estimated in a large generated data set by taking means of $y$ in bins of $r$.
This relationship is shown in \figurename~\ref{fig:regression} with solid red circles.
The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figurename~\ref{fig:regression} with open black squares.
These two relationships are statistically consistent.
The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black X-shaped symbols.
The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.
Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.

>>>>

% To do:
% ------
% - Change it to be about PHYSICS rather than natural science?
% - Fill in the outline.
% - Publish on the arXiv.
% - format for ICML Position Paper and submit.
% - Wait for the call from Stockholm.

@@

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
  We also show that the there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when expressive models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
  These remarks are intended to apply to all of the natural sciences, but most of the specific examples are in the astrophysics domain.
\end{abstract}

@@

\paragraph{Our contributions:}
\begin{itemize}
  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines. One is a confirmation bias that arises when simulations are replaced or augmented by emulators. Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses.
  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}. Many of these places are in the operational parts of scientific projects.
  \item Further, we demonstrate that there are contexts inside a natural-scientific analysis in which using the most expressive ML method delivers \emph{the most conservative approach} to the scientific problem. An example is in causal contexts, where providing immense flexibility to (say) a confounder model can strengthen conclusions about the causal channels of interest.
\end{itemize}

@@

\section{Why do we need machine learning in the natural sciences?}
In this \sectionname{} we give some idea of how and why ML has had such a big impact in the natural sciences, and how and why many scientific projects require ML technologies.
In most cases, the ML is required on the engineering or execution side of projects that are large-scale in some sense.
\begin{description}
  \item[Label transfer:] Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
  In this case, if it is extremely expensive to label more objects, a regression can be trained on the few labeled data points and then the trained regression used to label all the rest.
  Below (\sectionname~\ref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
  \item[Classification:] 
  \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
  The most extreme examples of this are in particle physics, where detectors (like \textsl{Atlas} \cite{atlas}) must decide whether to trigger a data-saving event in a tiny fraction of second.
  It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
  \item[Speeding up simulations:] In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
  These simulations tend to be very computationally expensive, since they often span large ranges of spatial or time scales.
  ML regressions can be trained to emulate the simulations, or patch up low-resolution simulations to higher resolutions.
  HOGG: THERE ARE MANY different approaches to emulation; summarize here?
  Below (\sectionname~\ref{sec:bad}) we argue that the introduction of ML-based emulators can introduce unwanted confirmation bias.
  \item[Modeling nuisances:] In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \sectionname~\ref{sec:good}) even conservative.
  \item[Outlier detection:] In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
  A recent example in astronomy is the discovery of fast radio bursts \cite{grbs}, which started as nuisances in imaging projects and have turned out to be interesting astrophysical objects.
  Since expressive unsupervised ML methods can describe accurately a complex distribution of data, they can also be used to find rare data points that are unlike any elements of the training data.
  \item[Making discoveries?] There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
  There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1, foundation2}.
\end{description}
The short summary is that we need ML in the natural sciences; we can't live without it.
The question here is how we will use it safely?

@@

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases.
Likelihoods can produce unbiased point estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimates.
The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace regressions with generative models that can produce likelihoods HOGG CITE THINGS.

@@

HOGG: DO WE CLEARLY SAY: ML estimates are always biased, maximum-likelihood estimates sometimes aren't. That's important if you are releasing data for downstream re-use.

@@

\section{When is machine learning good for natural science?}\label{sec:good}
HOGG: BE SURE to say what is in common across all of these examples. They are joined by a thread!

\paragraph{Execution and operations:}

@@

\clearpage\appendix
\section{Population-level biases in regression outputs}\label{app:toy}
A machine-learning (ML) regression learns the best set of parameters (weights and thresholds, for example) $\hat{\theta}$ of an expressive function $f(x;\theta)$ from a training set of labeled data.
The learned function $f(x;\hat{\theta})$ delivers a predicted or estimated label $\hat{y}$ for any data point with a complete set of features $x$.
Thus an estimated label $\hat{y}_\ast$ for a data point with features $x_\ast$ has taken information \emph{both} from the features $x_\ast$ and \emph{also} from the features and labels of all the data points in the training set used to set the parameters $\hat{\theta}$.
In applications in which regression-estimated labels are used, it matters \emph{how much information} is coming from the individual data-point's features and how much from the original training set.

To illustrate these ideas, we construct a toy data analysis.
This toy will make a general point, but it has a decidedly astronomical feel to it.
The complete toy data set is generated by a process with the algorithm below.
This algorithm makes references to objects $\xi_\eta$ and $\xi_\zeta$ which are generated prior to the start.
\begin{hoggnumerate}
    \item A floating-point value of a known parameter called ``guiding radius'' $r$ is generated in the range $0<r<14$.
    \item A floating-point value of a latent parameter called ``age'' $\eta$ is generated from a Gaussian with mean $14 - r$ and variance 4.
    \item The point is discarded if the age is outside the range $0<\eta<14$. If the point is not discarded:
    \item A $K$-dimensional latent vector $\zeta$ with $K=14$ is generated from a Gaussian with zero mean and variance 1.
    \item A $M$-dimensional latent vector $\xi$ with $M=110$ is created by $\xi = \xi_\eta\,\eta + \xi_\zeta\cdot\zeta$, where $\xi_\eta$ is a $M$-vector and $\xi_\zeta$ is a $M\times K$ matrix.
    The elements of $\xi_\eta$ and $\xi_\zeta$ were drawn (before the start) from Gaussians with zero mean and variances $0.01/K$ and $1/K$ respectively.
    \item A rectified latent vector $\tilde{\xi}$ is created by taking $1 + \xi$ and rectifying all pixels to lie in the range $[0, 1]$.
    \item A $M$-dimensional feature vector $x$ called ``the data'' is generated from a Gaussian with mean $\tilde{\xi}$ and variance $0.0025$.
    \item A label $y$ called ``measured age'' is generated from a Gaussian with mean $\eta$ and variance 1.
\end{hoggnumerate}
The data generated this way are shown in the top panels of \figurename~\ref{fig:regression}.
The data contain a non-trivial relationship between label $y$ (measured age) $y$ and known parameter $r$ (guiding radius).
This relationship is shown in \figurename~\ref{fig:regression}.
It isn't precisely linear because of the removal of data points with latent ages $\eta$ outside the range $0<\eta<14$.

We construct a three-layer multilayer perceptron (MLP) (using the scikit-learn implementation) with layer sizes of 64, 32, and 16 neurons.
The model is trained on a training set of 4096 data points with features $x$ and labels $y$.
The trained model takes as input features $x$ and outputs estimated labels $\hat{y}$.
The model is validated on a validation set of 2048 data points with features $x$ and labels $y$.
The test set is a much larger set of $10^5$ data points with features $x$ and no labels.
The model is used to make estimated labels $\hat{y}$ in the full test set.
The validation and test labels are shown in the bottom panels of \figurename~\ref{fig:regression}.
Importantly, \emph{the training, validation, and test samples are drawn from exactly the same process}, with the same distribution in all properties.
\begin{figure}[p!]
\includegraphics[width=0.49\textwidth]{notebooks/data_examples.png}
\includegraphics[width=0.49\textwidth]{notebooks/training_data.png} \\
\includegraphics[width=0.49\textwidth]{notebooks/validation.png}
\includegraphics[width=0.49\textwidth]{notebooks/test_data_results.png}
\caption{\sffamily%
Visualization of the toy regression. \textsl{Top-left:} Random examples of the data vectors $x$, which are one-dimensional images generated from a linear model plus a nonlinearity created by two rectifications. Details of the data generation are given in the text. Each example $x$ is labeled on the right side by the value of its label $y$. \textsl{Top-right:} The training-set labels $y$, plotted against the known parameter $r$, which is not used in the regression (only the vectors $x$ are used). Also shown are solid red circles showing the true mean relationship between $y$ and $r$ in the toy data. Open black squares show the empirical mean relationship measured in bins in the training-set data. \textsl{Bottom-left:} Validation of the trained regression in the validation set, showing that the label estimates $\hat{y}$ are noisy (as expected given the problem set-up) but not strongly biased. \textsl{Bottom-right:} The regression estimates $\hat{y}$ in the very large test set, plotted against the known parameter $r$. Also shown are the same solid red circles and open black squares as in the top-right plot. Black X-shaped symbols show the mean relationship between $\hat{y}$ and $r$. The relationship shown by the Xs is very precise but biased far away from the true relationship, unlike the relationship shown by the open squares (measured in the training set alone).\label{fig:regression}}
\end{figure}

The regression-estimated labels $\hat{y}$ are related to the measured labels $y$ in the held-out validation set with a linear relationship with unit slope and zero intercept.
The comparison to the validation data (bottom-left panel of \figurename~\ref{fig:regression}) shows that they are not precisely estimated, but there is no evidence for strong bias at the individual-object level.

The true relationship between label $y$ (measured age) and known parameter $r$ (guiding radius) is estimated in a large generated data set by taking means of $y$ in bins of $r$.
This relationship is shown in \figurename~\ref{fig:regression} with solid red circles.
The empirical relationship between label $y$ and known parameter $r$ in the training-set data is shown in \figurename~\ref{fig:regression} with open black squares.
These two relationships are statistically consistent, as expected.
The empirical relationship in the test-set data between the estimated label $\hat{y}$ and the known parameter $r$ in the test-set data is shown in \figurename~\ref{fig:regression} with black X-shaped symbols.
The test-set relationship is strongly biased, deviating from the true relationship at the edges of the $r$ range.

HOGG: TAKE A PARAGRAPH TO EXPLAIN THIS CLAIM: Given the large size of the test set, the precision of the result is high: This bias in the estimated relationship is larger than 30-sigma in the worst data points.
